External priors for the next generation of CMB experiments.

### 2015-12-07 Notes/N_eff_prior_paper_v3.tex

\date{\today}
\begin{abstract}
Planned cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, and dark energy. 
The low level of noise, together with improved angular resolution, will increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure. Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments. Here, we examine quantitatively the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 (S4) experiment on neutrino and dark energy properties.
We find that CMB S4 constraints on neutrino mass could be strongly enhanced by external constraints on the cold dark matter density $\Omega_{c}h^{2}$ and the Hubble constant $H_{0}$. If the largest scales ($\ell<50$) will not be measured, an external prior on the primordial amplitude $A_{s}$ or the optical depth $\tau$ will also be important. A CMB constraint on the number of relativistic degrees of freedom, $N_{\rm eff}$, will benefit from an external prior on the spectral index $n_{s}$ and the baryon energy density $\Omega_{b}h^{2}$. Finally, an external prior on $H_{0}$ will help constrain the dark energy equation of state ($w$).
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
Since their earliest incarnations, Cosmic Microwave Background (CMB) experiments have been crucial in furthering our understanding of the universe. They will maintain their role in the future, thanks to the unprecedented low level of noise and high resolution expected in the planned Stage IV (S4) experiment~\cite{2013arXiv1309.5383A}.
The recent past is  reassuring. For example, every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P, planck-collaboration:2014, 2003ApJS..148..175S, 2000ApJ...545L...5H,2000Natur.404..955D}. 
These led us from the first detection of anisotropy in the CMB temperature with COBE to a cosmic variance limited measurement of several acoustic peaks with Planck. With the improved sensitivity we have extended our understanding of the universe so that we now have solid evidence for a flat geometry and a well tested model $\Lambda$CDM with percent level constraints on its parameters.
The same success has characterized ground-based CMB experiments, where, e.g., the first detection of polarization anisotropy by DASI \cite{2002ApJ...568...38H} has been followed by a series of precise measurements of the damping tail and maps of the projected gravitational potential by, e.g., SPT~\cite{2011ApJ...743...28K} and ACT~\cite{2011ApJ...739...52D}.

The planned CMB-S4 experiment will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction. As has happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of several areas of astrophysics like dark matter, inflation, dark energy, and neutrinos.

The cosmological dependence on neutrinos is two-fold: the number of relativistic species $N_{\rm eff}$ in the early phase of the universe affects the damping tail of the CMB, and sum of the neutrino masses $\sum m_\nu$ affects the late-time growth of cosmic structure.
Because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe their number affects the expansion rate $H(z)$. This rate can be powerfully tested using the CMB, by carefully comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein). 
The future experiments, with their high resolution, will probe deep into the damping tale of the CMB power spectrum and will be sensitive to small variations from the canonical, 3-active neutrino prediction of $N_{\rm eff}=3.046$. The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($\sum m_{\nu}<230$ meV from \cite{planck-collaboration:2014}), neutrinos are still relativistic at the last scattering surface. However, massive neutrinos alter the growth of the large scale structure responsible for lensing of the CMB photons. Different neutrino masses consequently lead to different CMB lensing spectra. CMB-S4 will measure small scale temperature and polarization anisotropies with low noise, dramatically improving the lensing reconstruction. This will turn into a  precise constraints of the sum of neutrino masses with a possible hint of the hierarchies of the individual masses. %Quantitatively, the CMB alone will constrain the number of relativistic degree of freedom $N_{\rm eff}$ with $1\%$ precision and the total mass of neutrinos at $60\%$ with a factor of two improvement on $M_{\nu}$ when priors from baryon acoustic oscillations (BAO) data are introduced \cite{wu:2014}. The future generation of CMB experiments will be a big step in the understanding of the neutrino sector.

The CMB is sensitive also to the properties of dark energy (see, e.g., \cite{2010MNRAS.405.2639J}). Thanks to the current generation of experiments we know with extraordinary accuracy its energy density. The challenge for the next generation is to reveal the nature of this mysterious component. For example, a crucial step to identifying the mechanism driving cosmic acceleration will be to investigate any possible deviations in the equation of state, the ratio of pressure and energy density, from the value $w=-1$ predicted by a cosmological constant. 
Dark energy affects the CMB because it alters the universe's expansion and it consequently changes the distance to the last scattering surface. Furthermore different dark energy models lead to different rates of growth of large scale structure which are tested by CMB lensing. 
%Thanks to this sensitivity to cosmological structures present all the way from us to the last scattering surface, the CMB will also be a powerful probe of any time dependence of the dark energy equation of state.
However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$. As with the neutrino sector, dark energy constraints from the CMB 
will be improved by external experiments.

In the case of both neutrinos and dark energy then, an important lingering question is: What external information can be used to improve projected constraints by breaking degeneracies?
%The crucial importance of feedback from external experiments is true not only for the dark energy and neutrino sector. A general thrust in even the current generation of surveys is to combine information from different observations. %In this regard CMB makes no exception: external priors will be fundamental to improve the already tight CMB constraints on cosmological parameters. 
%In particular CMB will benefit from experiments like large scale structure clustering and weak lensing, BAO targeted experiments, and supernovae. 
Here we study the dependance of projected constraints from CMB on the external priors assumed.  For all the parameters of the standard $\Lambda$CDM with massive neutrinos ($\Omega_ch^2, \Omega_bh^2, A_s, n_s, \tau, H_0, \sum m_\nu$) together with extensions ($N_{\rm eff}$ and $w$) to the neutrino and dark energy sectors, we quantify the CMB constraints as a function of external priors. This extends the work of \citet{wu:2014}, which worked with a few fixed external priors, by quantifying the extent to which external information will improve the constraining power of a CMB-S4 experiment.

@@

For CMB experiments the Fisher matrix is related to the angular power spectra $\boldsymbol{C}_\ell$ by:
\begin{equation}
 F_{ij} = \sum_\ell \frac{2\ell+1}{2} f_{\rm sky} {\rm Tr} \left(  \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_i} \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_j}  \right)
 \label{eqn:Fij_def2}
 \end{equation}
 where $f_{\rm sky}$, the fraction of sky covered, is set to 0.75 throughout.
In this work we constrain cosmological parameters with CMB temperature and E-mode polarization together with the reconstructed lensing potential of large scale structure. Therefore, $\boldsymbol{C}_\ell$ in \refeq{Fij_def2} is:
 \begin{eqnarray}
 	\centering
		\mathbf{C}_\ell \equiv \left( \begin{array}{ccc}C_\ell^{TT} + N_\ell^{TT} & C_\ell^{TE} & C_\ell^{T\phi} \\ C_\ell^{TE} & C_\ell^{EE} + N_\ell^{EE} & 0 \\ C_\ell^{T\phi} & 0 & C_\ell^{\phi} + N_\ell^{\phi}\end{array}\right).
	\label{eqn:cov_definition}
\end{eqnarray}
The terms $N_\ell^{X}$ represent the instrumental noise power of the specific experiment and will be described at the end of this section.
Note that we are neglecting the term $C_\ell^{E\phi}$ since it  contains very little information while adding possible numerical issues~\cite{wu:2014,2013PhRvD..87h3008H}.
Furthermore, as in \cite{wu:2014}, we use \textit{unlensed} spectra and gaussian covariances in \refeq{Fij_def2}. 
Introducing eventual imperfect reconstruction of the unlensed spectra (delensing) and non-gaussian effects described in \cite{benoit-levy:2012} will go in the direction of degrading the parameters constrained derived in this work.
However, we expect this approximations to have a small impact on the final results and, for this reason, a more careful modeling is beyond the scope of this work. For example, we checked that using lensed spectra will change the final constraint by less than 10$\%$.

@@

We compute the power spectra $C_{\ell}$ and the noise power $N_{\ell}$ in \refeq{Fij_def2} using CAMB \cite{Howlett:2012mh,Lewis:1999bs} and
the derivatives in \refeq{Fij_def2} using a 5 point finite difference formula:
this high order approach allows us to use larger step-sizes around the fiducial parameters to compute derivatives. As a consequence the differences of power spectra corresponding to different values of the parameters are big enough to ensure numerical accuracy. We also test the robustness of this calculation by varying the derivative steps in the range $2-7\%$ . The constraints change by at most $10\%$ that should then be considered as a conservative estimate of numerical uncertainties. 
Our philosophy is to consider as few extensions as possible. Given the current success of the standard models of particle physics and cosmology, one of the primary goals of CMB-S4 will be to find cracks in these models. As such, the key question is: What information is needed to reliably conclude that an additional parameter is required. The natural first baby step is to include massive neutrinos, which (i) are a small extension to the Standard Model and (ii) are known to exist. So our fiducial cosmology is flat $\nu \Lambda$CDM, with assumed 
parameters from Table 2 of the \textit{Planck} best fit \cite{planck-collaboration:2014g}, i.e. $\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, $H_0 = 67.11$ km/s/Mpc, supplemented by an arbitrary choice of $\sum m_\nu$ $\simeq$ 85\ meV.  
We then extend the parameter space by introducing $N_{\rm eff}$ and $w$ as free parameters, in each case keeping the other parameter fixed.
The fiducial values of these are $N_{\rm eff}=3.046$ and a cosmological constant equation of state, $w=-1$. In this same spirit, we keep the curvature fixed to zero, as the level predicted in the vast majority of inflationary models is of order $10^{-5}$, much too small to have an impact on the parameters we consider. Similarly, we fix the helium abundance to its canonical value from Big Bang Nucleosynthesis, which depends only on the baryon density.
%It is important to notice that while we vary one parameter to compute derivatives in \refeq{Fij_def2} we keep all the others fixed with the exception of $\Omega_{\Lambda}$ which is always changed in order to keep the universe flat ($\Omega_{\rm k} = 0$).

The instrumental noise power $N_{\ell}^{T,E}$ and the lensing reconstruction $N_{\ell}^{\phi}$ in \refeq{Fij_def2} correspond to the optimistic level for CMB-S4 assumed in ~\cite{2013arXiv1309.5383A,wu:2014,2013PhRvD..87h3008H}.
For the temperature and E-mode polarization of the CMB, together with the improved depth and resolution we also assume that large scale foregrounds, like dust, are under control or negligible. This allows us to use all the polarization power spectrum multipoles out to $\ell_{\rm E,max}=5000$. We deal with the Poisson noise from point sources in the temperature signal by simply discarding all  small scale modes with $\ell>\ell_{\rm T,max}=3000$.
The remaining source of noise, the instrumental noise, is added to the power spectrum in the usual way:
 \begin{equation}
 	\centering
		N^{X}_\ell = s^{\, 2} \exp \left(\ell(\ell+1) \frac{\theta^{\ 2}_{\textsc{fwhm}}}{8\log2}\right),
	\label{eq:beamnoise}
\end{equation}
where $\theta^{\ 2}_{\textsc{fwhm}}$ is the FWHM of the experiment's beam and $s$ represents the instrumental white noise.
We use a level of noise $s = 0.58$ $\mu$K-arcmin for $X=T$ and a beam of $\theta_{\textsc{fwhm}}=1$ arcmin. This corresponds to the $N_{\rm det}=10^{6}$ case of \cite{wu:2014}. 
Note that the quoted noise in temperature and we assume that $s \rightarrow s\times \sqrt{2}$ in the case of polarization $ XX' = \{ EE, BB \}$. 
The noise $N_\ell^{\phi\phi}$ associated with the reconstructed $\phi$ spectrum is modeled assuming an iterative reconstruction technique \cite{seljak:2004}.

We also include information from current Baryon Acoustic Oscillation experiments and projected constraints from low-$l$ Planck polarization. 
As current BAO data, we follow \cite{allison:2015} and we take into account 6dFGRS \cite{beutler:2011}, SDSS MGS \cite{ross:2015} together with LOW-Z and C-MASS \cite{anderson:2014}.
The noise assumed for Planck corresponds to the ``Planck-pol'' specifications of \cite{allison:2015}, where noise levels were approximated by scaling the current sensitivities according to the Planck Blue Book.

@@

%==========
\begin{figure}[htbp]
\begin{center}
\includegraphics{{prior_omnuh2_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
\caption{
Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters. %($N_{det}=10^{6}$,$f_{sky}=0.75$, $2<\ell<5000$ ). 
Each prior is expressed in units of the error that would be obtained internally from the CMB experiment; e.g., $\sigma(H_0)_{\rm CMB}=0.5\%$, so the value of unity on the x-axis corresponds to imposing an external prior on $H_0$ of 0.6\%. $H_{0}$ and $\Omega_{c}h^{2}$ are shown to be the most important external priors to improve neutrino mass constraints. We chose a fiducial value $\sum m_{\nu}=85 $ meV and we impose a Planck-pol prior . The blue horizontal line correspond to the CMB S4+BAO constraint of \cite{wu:2014}.}
\label{fig:prior_omeganuh2}
\end{center}
\end{figure}
%==========

@@

The physical reason underlying this improvement is simple: lensing constrains $f_\nu$, the ratio of the neutrino density to the matter density. In order to extract the neutrino density (and therefore the neutrino mass), we need to know the matter density. A prior on  $H_{0}$ helps for a related, but subtle reason. 
%with this as well since the effects of external priors are not completely independent.
The position of the CMB peaks, well constrained by measurement, is a function $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014} so an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$, leading back to tighter neutrino mass constraints.
As found in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}, the projected error falls below 20 meV with these priors. 
The left-hand side of \reffig{prior_omeganuh2} shows that even tighter external constraints on either $H_0$ or $\Omega_ch^2$ would significantly improve the neutrino mass constraints.
%\yp{Alessandro, are you planning on including the $\Omega_b h^2$ discussion here?}

@@

Indeed, as recently pointed out by \cite{allison:2015}, a degeneracy exists between $\sum m_\nu$ and $A_{s}$ and, as a consequence, between $\sum m_\nu$ and $\tau$. This can be understood as follows.
The CMB lensing reconstruction is noisy at scales larger than the neutrino free streaming scale (see Fig. 5 of \cite{2013arXiv1309.5383A}). For this reason it is hard to measure the unsuppressed CMB lensing power spectra that is then compared to the small scale suppressed one to constrain the values of $f_{\nu}$. The unsuppressed amplitude of the lensing spectrum is actually better constrained by the primordial amplitude $A_{s}$, which is probed by the primordial CMB. Unfortunately, the CMB spectra are sensitive to $A_{s}\exp^{-2\tau}$, so 
a measurement of the optical depth is required to infer $A_{s}$ and therefore to tighten the constraints on $\sum m_\nu$. If CMB S4 will not be designed to measure scales $\ell<50$ an external prior on the optical depth from future 21 cm surveys \cite{liu:2015} or CMB satellite experiments like PIXIE \cite{kogut:2011} will be needed to fully exploit the CMB constraining power.

@@

In the standard model of cosmology, three active neutrinos are thermally produced in the early universe. Were they to decouple well before the epoch of electron-positron annihilation, their energy density after their decoupling would be equal to $3\times (7/8)\times (4/11)^{4/3}\,\rho_{\rm cmb}$, with the first factor capturing the contributions from the 3 active species; the second the difference between fermions and bosons; and the last the relative heating of the photons in the CMB by electron-positron annihilation. However, decoupling is not a discrete event and occurs close to the time of electron-positron annihilation, so the neutrinos share a bit in the heating, with the factor of 3 replaced by $N_{\rm eff}=3.046$. The additional fraction depends not only on well-known neutrino scattering rates but also on finite temperature quantum corrections. 
Upcoming experiments have the potential to measure this tiny deviation of $N_{\rm eff}$ from 3. This will be not only a precision test of our understanding of the Universe when it was about a second old. It also will constrain Beyond the Standard Model physics. For example, as pointed out in \cite{baumann:2015}, tight constraints on $N_{\rm eff}$ will allow us to rule out new particles with couplings that enabled them to thermalize early on but that decoupled when the temperature was above $\sim$100 GeV.
%decoupled early in the thermal history.

@@

The sensitivity to relativistic degrees of freedom comes from the effect of extra species on the damping tail of the CMB anisotropies \cite{2013PhRvD..87h3008H}, both in temperature and polarization, so parameters that also strongly affect this part of the spectrum, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are the most degenerate with $N_{\rm eff}$. As such, \reffig{prior_massless_neutrinos} shows that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to ensure a 3-sigma detection of the partial decoupling prediction. The bottom-most curve shows that if all parameters were constrained externally, then the projected error on $N_{\rm eff}$ would improve significantly. %\yp{The last sentence sounds a bit vague.}

>>>>

\date{\today}
\begin{abstract}
Planned cosmic microwave background (CMB) experiments can dramatically improve what we know about neutrino physics, inflation, and dark energy. 
Indeed the low level of noise, together with improved angular resolution, will increase the signal to noise of the CMB polarized signal as well as the reconstructed lensing potential of high redshift large scale structure. Projected constraints on cosmological parameters are extremely tight, but these can be improved even further with information from external experiments. Here, we examine quantitatively the extent to which external priors can lead to improvement in projected constraints from a CMB-Stage 4 (S4) experiment on neutrino and dark energy properties.
We find that CMB S4 constraints on neutrino mass could be strongly enhanced by external constraints on the cold dark matter density $\Omega_{c}h^{2}$ and the Hubble constant $H_{0}$. If the largest scales ($\ell<50$) will not be measured, an external prior on the primordial amplitude $A_{s}$ or the optical depth $\tau$ will also be important. Furthermore a CMB constraint on the number of relativistic degrees of freedom, $N_{\rm eff}$, will benefit from an external prior on the spectral index $n_{s}$ and the baryon energy density $\Omega_{b}h^{2}$. Finally, an external prior on $H_{0}$ will help constrain the dark energy equation of state ($w$).
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
Since their earliest incarnations, Cosmic Microwave Background (CMB) experiments have been crucial in furthering our understanding of the universe. They will maintain their role in the future, thanks to the unprecedented low level of noise and high resolution expected in the planned Stage IV (S4) experiment~\cite{2013arXiv1309.5383A}.
The recent past is reassuring. For example, every new generation of satellite experiments improved the level of sensitivity by almost a factor of ten compared to its predecessor, from the first generation instrument COBE to WMAP all the way to the current state of the art represented by Planck \cite{2015arXiv150201589P, planck-collaboration:2014, 2003ApJS..148..175S, 2000ApJ...545L...5H,2000Natur.404..955D}. 
These led us from the first detection of anisotropy in the CMB temperature with COBE to a cosmic variance limited measurement of several acoustic peaks with Planck. With the improved sensitivity we have extended our understanding of the universe so that we now have solid evidence for a flat geometry and a well tested model $\Lambda$CDM with percent level constraints on its parameters.
The same success has characterized ground-based CMB experiments, where, e.g., the first detection of polarization anisotropy by DASI \cite{2002ApJ...568...38H} has been followed by a series of precise measurements of the damping tail and maps of the projected gravitational potential by, e.g., SPT~\cite{2011ApJ...743...28K} and ACT~\cite{2011ApJ...739...52D}.
The planned CMB-S4 experiment will potentially measure the E-mode polarization with cosmic variance limited precision together with an order of magnitude improvement in B-mode measurement and lensing reconstruction. As has happened in the past, this new sensitivity together with the progress in the measurement of other cosmological probes will improve our understanding of several areas of astrophysics like dark matter, inflation, dark energy, and neutrinos.

The cosmological dependence on neutrinos is two-fold: the number of relativistic species $N_{\rm eff}$ in the early phase of the universe affects the damping tail of the CMB, and sum of the neutrino masses $\sum m_\nu$ affects the late-time growth of cosmic structure.
Because relativistic species, like neutrinos, are the main drivers of the cosmic expansion in the early Universe, their number affects the expansion rate $H(z)$. This rate can be powerfully tested using the CMB, by carefully comparing the sound horizon scale, obtained from the CMB peaks positions, and the Silk damping scale (see \cite{2013arXiv1309.5383A}, \cite{2013PhRvD..87h3008H} and references therein). 
The future experiments, with their high resolution, will probe deep into the damping tale of the CMB power spectrum and will be sensitive to small variations from the canonical, 3-active neutrino prediction of $N_{\rm eff}=3.046$. The total mass of neutrinos, on the other hand, has a modest effect on the CMB because, for the range of masses allowed by recent constraints ($\sum m_{\nu}<230$ meV from \cite{planck-collaboration:2014}), neutrinos are still relativistic at the last scattering surface. However, massive neutrinos alter the growth of the large scale structure responsible for lensing of the CMB photons. Different neutrino masses consequently lead to different CMB lensing spectra. CMB-S4 will measure small scale temperature and polarization anisotropies with low noise, dramatically improving the lensing reconstruction. This will turn into a precise constraints of the sum of neutrino masses with a possible hint of the hierarchies of the individual masses.

The CMB is sensitive also to the properties of dark energy (see, e.g., \cite{2010MNRAS.405.2639J}). Thanks to the current generation of experiments we know with extraordinary accuracy its energy density. The challenge for the next generation is to reveal the nature of this mysterious component. For example, a crucial step to identifying the mechanism driving cosmic acceleration will be to investigate any possible deviations in the equation of state, the ratio of pressure and energy density, from the value $w=-1$ predicted by a cosmological constant. 
Dark energy affects the CMB because it alters the universe's expansion and it consequently changes the distance to the last scattering surface. Furthermore different dark energy models lead to different rates of growth of large scale structure which are tested by CMB lensing. 
However, dark energy properties are strongly degenerate with other geometrical parameters like $H_{0}$ and $\Omega_{k}$. As with the neutrino sector, dark energy constraints from the CMB 
will be improved by external experiments.

In the case of both neutrinos and dark energy then, an important lingering question is: What external information can be used to improve projected constraints by breaking degeneracies?
Here we study the dependance of projected constraints from CMB on the external priors assumed. For all the parameters of the standard $\Lambda$CDM with massive neutrinos ($\Omega_ch^2, \Omega_bh^2, A_s, n_s, \tau, H_0, \sum m_\nu$) together with extensions ($N_{\rm eff}$ and $w$) to the neutrino and dark energy sectors, we quantify the CMB constraints as a function of external priors. This extends the work of \citet{wu:2014}, which worked with a few fixed external priors, by quantifying the extent to which external information will improve the constraining power of a CMB-S4 experiment.

@@

For CMB experiments the Fisher matrix is related to the angular power spectra $\boldsymbol{C}_\ell$ by:
\begin{equation}
 F_{ij} = \sum_\ell \frac{2\ell+1}{2} f_{\rm sky} {\rm Tr} \left(  \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_i} \boldsymbol{C}^{-1}_\ell( \theta) \frac{\partial \boldsymbol{C}_\ell}{\partial \theta_j}  \right)
 \label{eqn:Fij_def2}
 \end{equation}
 where $f_{\rm sky}$, the fraction of sky covered, is set to 0.75 throughout.
In this work we constrain cosmological parameters with CMB temperature and E-mode polarization together with the reconstructed lensing potential of large scale structure. Therefore, $\boldsymbol{C}_\ell$ in \refeq{Fij_def2} is:
 \begin{eqnarray}
 	\centering
		\mathbf{C}_\ell \equiv \left( \begin{array}{ccc}C_\ell^{TT} + N_\ell^{TT} & C_\ell^{TE} & C_\ell^{T\phi} \\ C_\ell^{TE} & C_\ell^{EE} + N_\ell^{EE} & 0 \\ C_\ell^{T\phi} & 0 & C_\ell^{\phi} + N_\ell^{\phi}\end{array}\right).
	\label{eqn:cov_definition}
\end{eqnarray}
The terms $N_\ell^{X}$ represent the instrumental noise power of the specific experiment and will be described at the end of this section.
Note that we are neglecting the term $C_\ell^{E\phi}$ since it contains very little information while adding possible numerical issues~\cite{wu:2014,2013PhRvD..87h3008H}.
Furthermore, as in \cite{wu:2014}, we use \textit{unlensed} spectra and gaussian covariances in \refeq{Fij_def2}. 
This means that we neglect eventual imperfect reconstruction of the unlensed spectra (delensing) and the non-gaussian effects described in \cite{benoit-levy:2012} .
However, we expect this approximations to have a small impact on the final results and, for this reason, a more careful modeling is beyond the scope of this work. For example, we checked that using lensed spectra will change the final constraint by less than 10$\%$.

@@

We compute the power spectra $C_{\ell}$ in \refeq{Fij_def2} using CAMB \cite{Howlett:2012mh,Lewis:1999bs} and
the derivatives in \refeq{Fij_def2} using a 5 point finite difference formula:
this high order approach allows us to use larger step-sizes around the fiducial parameters to compute derivatives. As a consequence the differences of power spectra corresponding to different values of the parameters are big enough to ensure numerical accuracy. We also test the robustness of this calculation by varying the derivative steps in the range $2-7\%$ . The constraints change by at most $10\%$ that should then be considered as a conservative estimate of numerical uncertainties. 
Our philosophy is to consider as few extensions as possible. Given the current success of the standard models of particle physics and cosmology, one of the primary goals of CMB-S4 will be to find cracks in these models. As such, the key question is: What information is needed to reliably conclude that an additional parameter is required. The natural first baby step is to include massive neutrinos, which (i) are a small extension to the Standard Model and (ii) are known to exist. So our fiducial cosmology is flat $\nu \Lambda$CDM, with assumed 
parameters from Table 2 of the \textit{Planck} best fit \cite{planck-collaboration:2014g}, i.e. $\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, $H_0 = 67.11$ km/s/Mpc, supplemented by an choice of $\sum m_\nu$ $\simeq$ 85\ meV to be consistent with \cite{wu:2014}.  
We then extend the parameter space by introducing $N_{\rm eff}$ and $w$ as free parameters, in each case keeping the other parameter fixed.
The fiducial values of these are $N_{\rm eff}=3.046$ and a cosmological constant equation of state, $w=-1$. In this same spirit, we keep the curvature fixed to zero, as the level predicted in the vast majority of inflationary models is of order $10^{-5}$, much too small to have an impact on the parameters we consider. Similarly, we fix the helium abundance to its canonical value from Big Bang Nucleosynthesis, which depends only on the baryon density.

The instrumental noise power $N_{\ell}^{T,E}$ and the lensing reconstruction $N_{\ell}^{\phi}$ in \refeq{Fij_def2} correspond to the optimistic level for CMB-S4 assumed in ~\cite{2013arXiv1309.5383A,wu:2014,2013PhRvD..87h3008H}.
For the temperature and E-mode polarization of the CMB, together with the improved depth and resolution we also assume that large scale foregrounds, like dust, are under control or negligible. This allows us to use all the polarization power spectrum multipoles out to $\ell_{\rm E,max}=4500$. We deal with the Poisson noise from point sources in the temperature signal by simply discarding all small scale modes with $\ell>\ell_{\rm T,max}=3000$.
The remaining source of noise, the instrumental noise, is added to the power spectrum in the usual way:
 \begin{equation}
 	\centering
		N^{X}_\ell = s^{\, 2} \exp \left(\ell(\ell+1) \frac{\theta^{\ 2}_{\textsc{fwhm}}}{8\log2}\right),
	\label{eq:beamnoise}
\end{equation}
where $\theta^{\ 2}_{\textsc{fwhm}}$ is the FWHM of the experiment's beam and $s$ represents the instrumental white noise.
We use a level of noise $s = 0.58$ $\mu$K-arcmin for $X=T$ and a beam of $\theta_{\textsc{fwhm}}=1$ arcmin. This corresponds to the $N_{\rm det}=10^{6}$ case of \cite{wu:2014}. 
Note that the quoted noise in temperature and we assume that $s \rightarrow s\times \sqrt{2}$ in the case of polarization $ XX' = \{ EE, BB \}$. 
The noise $N_\ell^{\phi\phi}$ associated with the reconstructed $\phi$ spectrum is modeled assuming an iterative reconstruction technique \cite{seljak:2004}.

We also include information from current Baryon Acoustic Oscillation experiments and projected constraints from low-$l$ Planck polarization. 
As current BAO data (labeled BAO15), we follow \cite{allison:2015} and we take into account 6dFGRS \cite{beutler:2011}, SDSS MGS \cite{ross:2015} together with LOW-Z and C-MASS \cite{anderson:2014}.
The noise assumed for Planck corresponds to the ``Planck-pol'' specifications of \cite{allison:2015}, where noise levels were approximated by scaling the current sensitivities according to the Planck Blue Book.

@@

%==========
\begin{figure}[htbp]
\begin{center}
\includegraphics{{prior_omnuh2_snow_mass_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
\caption{
Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters. 
Each prior is expressed in units of the error that would be obtained internally from the CMB S4 experiment (+Planck+current BAO); e.g., $\sigma(H_0)_{\rm S4 + Planck Pol + BAO15}=0.6\%$, so the value of unity on the x-axis corresponds to imposing an external prior on $H_0$ of 0.6\%.
The horizonal line correspond to CMB S4 experiment + Planck Pol + DESI.
$H_{0}$ and $\Omega_{c}h^{2}$ are shown to be the most important external priors to improve neutrino mass constraints. We chose a fiducial value $\sum m_{\nu}=85 $ meV and we impose a Planck-pol prior . The blue horizontal line correspond to the CMB S4+BAO constraint of \cite{wu:2014}.}
\label{fig:prior_omeganuh2}
\end{center}
\end{figure}
%==========

@@

The physical reason underlying this improvement is simple: lensing constrains $f_\nu$, the ratio of the neutrino density to the matter density. In order to extract the neutrino density (and therefore the neutrino mass), we need to know the matter density. A prior on $H_{0}$ helps for a related, but subtle reason. 
The position of the CMB peaks, well constrained by measurement, in a flat universe is a function $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014} 
. For this reason an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$, leading back to tighter neutrino mass constraints.
As found in \cite{2013arXiv1309.5383A,pan:2015,allison:2015}, the projected error falls below 20 meV with these priors. 
The left-hand side of \reffig{prior_omeganuh2} shows that even tighter external constraints on either $H_0$ or $\Omega_ch^2$ would significantly improve the neutrino mass constraints.

@@

Indeed, as recently pointed out by \cite{allison:2015}, a degeneracy exists between $\sum m_\nu$ and $A_{s}$ and, as a consequence, between $\sum m_\nu$ and $\tau$. This can be understood as follows.
The CMB lensing reconstruction is noisy at scales larger than the neutrino free streaming scale (see Fig. 5 of \cite{2013arXiv1309.5383A}). For this reason it is hard to measure the unsuppressed CMB lensing power spectra that is then compared to the small scale suppressed one to constrain the values of $f_{\nu}$. The unsuppressed amplitude of the lensing spectrum is actually better constrained by the primordial amplitude $A_{s}$, which is probed by the primordial CMB. Unfortunately, the CMB spectra are sensitive to $A_{s}\exp^{-2\tau}$, so 
a measurement of the optical depth is required to infer $A_{s}$ and therefore to tighten the constraints on $\sum m_\nu$. If CMB S4 will not be designed to measure scales $\ell<50$ an external prior on the optical depth from future 21 cm surveys \cite{liu:2015} or CMB satellite experiments like PIXIE \cite{kogut:2011} or balloon like SPIDER \cite{filippini:2010}, will be needed to fully exploit the CMB constraining power.

@@

In the standard model of cosmology, three active neutrinos are thermally produced in the early universe. Were they to decouple well before the epoch of electron-positron annihilation, their energy density after their decoupling would be equal to $3\times (7/8)\times (4/11)^{4/3}\,\rho_{\rm cmb}$, with the first factor capturing the contributions from the 3 active species; the second the difference between fermions and bosons; and the last the relative heating of the photons in the CMB by electron-positron annihilation. However, decoupling is not a discrete event and occurs close to the time of electron-positron annihilation, so the neutrinos share a bit in the heating, with the factor of 3 replaced by $N_{\rm eff}=3.046$. The additional fraction depends not only on well-known neutrino scattering rates but also on finite temperature quantum corrections. 
Upcoming experiments have the potential to measure this tiny deviation of $N_{\rm eff}$ from 3. This will be a precision test of our understanding of the Universe when it was about a second old. It also will constrain Beyond the Standard Model physics. For example, as pointed out in \cite{baumann:2015}, tight constraints on $N_{\rm eff}$ will allow us to rule out new particles with couplings that enabled them to thermalize early on but that decoupled when the temperature was above $\sim$100 GeV.

@@

The sensitivity to relativistic degrees of freedom comes from the effect of extra species on the damping tail of the CMB anisotropies \cite{2013PhRvD..87h3008H}, both in temperature and polarization, so parameters that also strongly affect this part of the spectrum, like the slope $n_s$ and the baryon density $\Omega_bh^2$, are the most relevant to improce $N_{\rm eff}$ constraint. As such, \reffig{prior_massless_neutrinos} shows that obtaining external priors on either of these would reduce the errors on $N_{\rm eff}$ to ensure a 3-sigma detection of the partial decoupling prediction. The bottom-most curve shows that if all parameters were constrained externally, then the projected error on $N_{\rm eff}$ would improve significantly.
